### 2021-12-11 manuscript.tex
Fix small configs [ci skip]
This build is based on
https://github.com/lubianat/quali_phd/commit/52efe9a95b60519392be126b6111a6ac8f1ea5da.

This commit was created by the following CI build and job:
https://github.com/lubianat/quali_phd/commit/52efe9a95b60519392be126b6111a6ac8f1ea5da/checks
https://github.com/lubianat/quali_phd/actions/runs/1567904151

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/bd9740705b22bf1d745d859aebf5f1b9fc80b3e1/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/bd9740705b22bf1d745d859aebf5f1b9fc80b3e1}{lubianat/quali\_phd@bd97407}
on December 10, 2021.

@@

Human Cell Atlas publications (https://www.humancellatlas.org/publications) were selected and abstracts were annotated as richly as possible with Wikidata IDs using the hypothes.is annotation system (https://web.hypothes.is/). One article {[}\protect\hyperlink{ref-1GmbExweg}{2}{]}, describing the complete Human Cell Atlas project, was annotated in full. Annotations were retrieved via the hypothes.is API and processed with custom Python and R scripts (https://github.com/lubianat/ann/tree/main/hypothesis\_parsing).

@@

As part of the research-action process, I have joined the Cell Ontology working group.
I participate in the monthly meetings and sporadic workshops, learning and contributing to the discussions.
Additionally, I contribute to the ontology development, actively engaging in the Cell Ontology GitHub repository (\url{https://github.com/obophenotype/cell-ontology}) and contributing with new terms and assertions.
I edit the ontology with the software for ontology editing Protégé v. 5.5.0 {[}\protect\hyperlink{ref-CHsAoEvr}{124}{]}.

@@

We targeted the question: which cell type definition allows crafting coherent biological statements?
The goal was not to say what cell types \emph{are}, but what they can be for a consistent representation on an ontology or a knowledge graph, like Wikidata.
We avoided the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{125}{]} for an example).
Even though such a distinction is essential for theoretical research, it is not required to represent biomedical experiments formally.

In a preprint derived from this PhD project {[}\protect\hyperlink{ref-uCE45Jh8}{126}{]}, we proposed naming conventions for different cell types classes.
Much of the literature mixes cell types in one species (e.g., when dealing with a cell type as an evolutionary unit) or multiple species (e.g., in the Cell Ontology).
It is helpful to distil these different concepts into names. Given the importance of the species' concept in biological classification {[}\protect\hyperlink{ref-AHGRV6KQ}{127}{]}, we derive a species-centric view on naming classes of cell types.
The four classes (Figure \ref{fig:classes}) we propose are as follows:

@@

Refining the different concepts around the notion of ``cell type'' is important but will require decades before a reasonable consensus. Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as any class of cells described by a domain expert with evidence of the reality of its instances.
The requirement of evidence of existence in the material world is based on the principle of instantiation of ontological realism {[}\protect\hyperlink{ref-5ZFRyTy5}{128}{]}. Barry Smith and Werner Ceuster state ``A term should be included in a reference ontology only if there is experimental evidence that instances to which that term refers exist in reality. (`Exists' here should be understood in a tenseless sense in order to accommodate, for example, universals pertaining to extinct species as well as universals such as swarm or hurricane which are instantiated only intermittently.)''
Thus, in this work, one minimum requirement for a cell type to be catalogued is a public description of the class, with evidence for the existence of instances of the class in reality.

By ``class'', we mean an abstract entity in a sense intended by the multilevel theory (MLT) of conceptual modelling {[}\protect\hyperlink{ref-eP47FILn}{129}{]}
Figure \ref{fig:multilevel} displays a simplified version of MLT adopted throughout this project.
In this framework, real-world entities are materializations of \emph{individuals}.
\emph{Individuals} are theoretical constructs that are (1) thought to exist or have existed, as per the principle of instantiation, and (2) refer to only one (01) material entity at any point in time.
For example, Wikidata has entries for people, e.g.~``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' are considered \emph{individuals} by Multi-Level Theory.
Other examples of \emph{individuals} include ``Albert Einstein's brain (\href{https://www.wikidata.org/wiki/Q2464312}{Q2464312})'' and the ``Christ the Redeemer statue (\href{https://www.wikidata.org/wiki/Q79961}{Q79961})''.

@@

For the practical purpose adopted here, we avoid the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{125}{]} for an example). We also consider only the cell as it was observed in an experiment, not necessarily the future conditions of any cell (i.e., the ``cell fate''). {[}\protect\hyperlink{ref-OjxQDZpx}{31}{]}\\
Even though such a distinction is an important topic for theoretical research, it is outside the initial scope of this work.

@@

We also note that we made a judgment call to use the term ``cell type'' to emphasize the focus on types as classes (or ``kinds'') in contrast to real-world objects.
The term ``cell class'' is also used in the literature and is a suitable synonym for our notion of cell type.
We opted to frame our work around the term ``cell type'' due to its historical usage and familiarity for the life sciences community.
Other related terms a ``cell set,'' ``cell population,'' and ``cell cluster,'' can also reminisce of a specific, countable group of cells, frequently from the same experiment.
The term ``cell identity'' has also been suggested for avoiding the cell type/cell state dilemma {[}\protect\hyperlink{ref-WKbly37M}{42}{]}, but we avoid it to emphasize a nominalistic perspective (in the Popperian sense{[}\protect\hyperlink{ref-1C6LI68h6}{130}{]}).
In doing so, we reinforce the intent to represent the cell types \_reported to exist\_instead of stating bluntly which cell types \emph{exist} or are \emph{essential} for human beings.

@@

Making the Human Cell Atlas more useful via Wikidata also includes the connection of related databases.
PanglaoDB {[}\protect\hyperlink{ref-M0lqRYJb}{131}{]} {[}\protect\hyperlink{ref-T3PlZ3Vh}{132}{]} is a publically-available database that contains data and metadata on hundreds of single-cell RNA sequencing experiments.
It provides extensive information on cell types, genes, tissues, and cell type markers, obtained via automatic and manual methods.
It also displays a rich web user interface for easy data acquisition, including database dumps for bulk downloads.

As of 8 December June 2021, the article describing PanglaoDB had been cited 230 times.
Despite its use by the community, the database is on a 3-star category for Linked Open Data {[}\protect\hyperlink{ref-OGXYtGZ8}{133}{]} as it does not use the open semantic standards from W3C (RDF and SPARQL) needed for a 4-star rank, neither the links to external data via standard identifiers that make datasets 5-star.
Improving the data format toward W3C's gold standards is a valuable step in making biological knowledge FAIR (Findable, Accessible, Interoperable, and Reusable).
Thus, we aimed to provide a case study of making the core information of PandlaoDB available in a 5-star Linked Open Data Format while improving the modelling of the necessary concepts on Wikidata.

As of August 2020, Wikidata had 264 items categorized as a ``cell type'', considerably less than in specialized cell catalogues, which count over two thousand cell types {[}\protect\hyperlink{ref-agnqfdk6}{49},\protect\hyperlink{ref-4AEy2xhQ}{134}{]}.
Strikingly, there were also 23 items categorized as instances of ``cell (Q7868)''.
This classification is imprecise, as an instance of cell would be an individual named cell from a single named individual, an example of conceptual disarray that often occurs on Wikidata. {[}\protect\hyperlink{ref-imSZTi52}{135}{]}

@@

Now that we re-formatted the markers on PanglaoDB as Linked Open Data, we can make queries that were not possible before, including federated queries with other biological databases, such as Uniprot {[}\protect\hyperlink{ref-HUJ9NzAb}{136}{]}
and Wikipathways {[}\protect\hyperlink{ref-AgOOCR24}{137}{]}.
Due to previous similar reconciliation projects, Wikidata already contains information about genes, including their relations to Gene Ontology (GO) terms.

@@

This work exemplifies the power of releasing Linked Open Data via Wikidata, and provides the biomedical community with the first semantically accessible, 5-star LOD dataset of cell markers, easily reachable from Wikidata's SPARQL Query Service (\url{https://query.wikidata.org/}).
Alongside other case studies of biomedical data integration to Wikidata (see {[}\protect\hyperlink{ref-mPoPwN77}{116}{]}, it contributes with tools and practices to serve as basis for contributors.)
The work also paves the way for reconciling of other databases for cell-type markers, such as CellMarker {[}\protect\hyperlink{ref-chGii6yw}{47}{]}, labome {[}\protect\hyperlink{ref-rhRRCtlA}{138}{]}, CellFinder {[}\protect\hyperlink{ref-4AEy2xhQ}{134}{]} and SHOGoiN/CELLPEDIA {[}\protect\hyperlink{ref-6uWWsiSq}{139}{]}) (if the owners give proper authorization).
The approach we took here can be applied to any knowledge set of public interest, providing a low-cost and low-barrier platform for sharing biocurated knowledge in gold-standard format.

@@

Reading scientific articles is an integral part of the routine of modern scientists.
Although several literature-management software are available {[}\protect\hyperlink{ref-SdbIYKUa}{140}{]}, the process of reading is mainly artisanal.
There are no standard guidelines on how to probe the literature organize notes for biomedical researchers.
Thus, while reading and studying is a core activity, there are few (if any) protocols for the efficient screening of scientific articles.

Other professional traditions have dealt with similar issues in the past.
Notetaking is vital to keep track of financial balances and avoid costly problems in accounting.
Double-entry bookkeeping was developed in the 13th century as a professional solution for notetaking in accounting where ``every entry to an account requires a corresponding and opposite entry to a different account.'' {[}\protect\hyperlink{ref-uYuz0opI}{24}, =Double-entry\_bookkeeping\&oldid=1055066428{]}
In software development, Test-Driven Development (TDD) is a popular methodology where tests for code snippets are written before the code itself, therefore ensuring that written software passes minimum quality standards.
The similarities of Double-entry bookkeeping and TDD are diverse {[}\protect\hyperlink{ref-KNGv5MPp}{141}{]}, but for our purpose, here suffices to see both as professionalized systems that promote better quality and accountability of works.

In the humanities, there is a well-established practice of annotations of readings.
The annotation skills are part of standard academic training in the humanities {[}\protect\hyperlink{ref-rPKBwmYh}{142}{]}{[}\protect\hyperlink{ref-url}{\textbf{url?}}{]}.
An influential work in presenting methods for academic reading in the humanities is Umberto Eco's book ``How to Write a Thesis'' {[}\protect\hyperlink{ref-1HBVPtZGp}{143}{]}, which outlines not only \emph{how} to annotate the literature that basis an academic thesis, but also \emph{why} to do so.
The book, written originally in 1977, is still influential today.
Still, its theoretical scope (roughly the humanities) and its date preceding the digital era limits the extent to which it applies to the biomedical sciences.

@@

The \texttt{docs/} directory contains the live dashboard from the readings, which will be discussed in the following sessions.
The \texttt{downloads/} directory hosts the pdfs of the articles read with the system.
These are not committed to the repository and are only stored locally.
The \texttt{notes/} directory contains markdown files, one for each article read.
The \texttt{src/} directory contains the python code with the system's mechanics.
They contain helper functions for the command line commands discussed below:
- \texttt{wread} which receives a Wikidata QID for an article and outputs (1) a notes document, (2) a pdf for the paper obtained from Unpaywall {[}\protect\hyperlink{ref-15luL9zZC}{144}{]} and (3) an updated version of the dashboard HTML files in the \texttt{docs/} directory.
- \texttt{pop}, which ``pops'' an article from \texttt{toread.md} and runs \texttt{wread} for it
- \texttt{wadd}, which takes an URL for a Wikidata SPARQL query and adds new QIDs to \texttt{toread.md}
- \texttt{wadd\_all}, which parses \texttt{config.yaml} for recurrent SPARQL queries and runs \texttt{wadd} for each
- \texttt{wlog}, which adds, commits and pushes recent readings and dashboard updates to GitHub

All the structures described so far are commonly shared by any user of Wikidata Bib.
To personalize the use of the system, the user edits three plain text files.
\texttt{toread.md} hosts plain text QIDs of the articles that will be read.
These can be added either manually or via wadd.
While the \texttt{pop} command only sees QIDs, articles titles or other identifiers can temporarily be added to \texttt{toread.md} without breaking the system.
\texttt{index.md} hosts a numbered list of topics of interest.
This file plays the role of Umberto Eco's work plan, with the topics of interest for the academic. {[}\protect\hyperlink{ref-1HBVPtZGp}{143}{]}
These are used to tag articles for retrieval in a later step.
\texttt{config.yaml} contains shortcuts for different reading lists.
This is better explained by example.
In my \texttt{to\ read.md} file there are two reading lists, one following a \texttt{\#\ Cell\ types} header and another following a \texttt{\#\ Biocuration} header.
My \texttt{config.yaml} contains the following snippet:

@@

The Wikidata bib framework is coupled with a discipline of daily reading.
The discipline is inspired by Robert Cecil Martin's description of Test Driven Development in the book ``Clean Code'', which includes not only a technical description but a \emph{school of thought} of how software development might be approached. {[}\protect\hyperlink{ref-13HqB23xH}{145}{]}
Every day, I read one article of each list, using the notetaking station displayed in Figure {[}\protect\hyperlink{ref-fig}{\textbf{fig?}}{]}: notetaking.
The constancy of reading allows steady coverage of the relevant literature.
While the discipline has worked for this research project, it is not required to use the Wikidata Bib system.

@@

The discipline also includes, whenever possible, an improvement of the metadata about the article on Wikidata.
In {[}\protect\hyperlink{ref-fig}{\textbf{fig?}}{]}: notetaking B are shown the links included in the dashboard.
A link to a Scholia {[}\protect\hyperlink{ref-hxzL9pmm}{106}{]} profile allows identification of related articles from a series of pre-made SPARQL queries probing bibliography data on Wikidata.
While Scholia provides an overview of a given article, it does not allow direct curation of the metadata.
For that, two links are provided, one to Wikidata and one to Author Disambiguator {[}\protect\hyperlink{ref-1A9RvszKC}{146}{]}.
By accessing the Wikidata page for the entity, one can add new triples, for example, curating authors and topics of the article, which are then used by Scholia and by Wikidata Bib's dashboard.
Author Disambiguator is a wrapper of an Wikimedia API that facilitates disambiguating author names to unique identifiers on Wikidata, thus feeding the public knowledge graph of publication and authors.\\
Finally, a link to the article's DOI or full-text URL is provided and serves as a fallback when the automatic download fails.
Of note, while the metadata curation has a technical benefit to Wikidata and the dashboard, it also plays a theoretical role.
By curating metadata on authors, the user of Wikidata Bib can better understand the people they read, and expand their metascientific perspective on their domain of interest.

@@

To organize the code for rendering the dashboard, we created a python package, wbib, and deposited it in PyPi, making it available via \texttt{pip}. {[}\protect\hyperlink{ref-6chnW6cc}{147}{]}.
The package implements the logic for rendering complex Wikidata-based academic dashboards and is available in GitHub at \url{https://github.com/lubianat/wbib}.
It allows the user to build dashboards based on Wikidata records of information such as gender of authors, the region of author's institutions, topics of articles and similar metascientific information.
The dashboard is composed of SPARQL queries written for the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{85}{]}
It also allows users to feed an arbitrary list of articles and obtain a custom dashboard.
Wikidata Bib obtains the HTML dashboards after feeding wbib the lists of articles read in total (\texttt{index.html}) or in pre-determined time spans (\texttt{last\_day.html}, \texttt{past\_week.html} and \texttt{past\_month.html} )

\begin{figure}
\hypertarget{fig:dashboard}{%
\centering
\includegraphics[width=0.85\textwidth,height=\textheight]{images/wikidata_bib_display.svg}
\caption{Wikidata Bib queries for institutions of authors and most-read venues}\label{fig:dashboard}
}
\end{figure}

The dashboard includes not only a basic list of read articles, but also statistics on most read authors and most-read venues.
It also displays an interactive map of the institutions of articles read, permitting a glance at geographic biases in activities.
An example of queries is shown in \ref{fig:dashboard}.
As the queries are rendered live, they evolve in quality with the growth of Wikidata.
Finally, the clean 5-star-open data format enables users to adapt the queries to include different aspects of Wikidata.
For example, table \ref{tbl:articles_read_hca} showcases 10 articles that (1) I have read in the past year and (2) were authored by a speaker of the 1st Human Cell Atlas Latin America Single Cell RNA-seqData Analysis Workshop {[}\protect\hyperlink{ref-1hag8XE6}{148}{]}.
One practical application that the dashboard enables, thus, is to identify people in an event, institution or location that the user has read before, therefore catalyzing the possibility of collaborations.
Anecdotally, this strategy was tested successfully at Biohackathon Europe 2021 {[}\protect\hyperlink{ref-kHL3NVxk}{149}{]}, where I used the system both to identify possible collaborators and as a conversation starter.

@@

To start the improvement in the interplay of both databases, we proposed and got the approval of a specific Wikidata identifier for the Cell Ontology, the ``Cell Ontology ID'' (\url{https://www.wikidata.org/wiki/Property:P7963}).
IDs can be added to Wikidata entities and connected them to external databases enabling integrative SPARQL queries.
Besides using the common Wikidata interface, one can crowd-curate identifiers via a 3rd-party service, Mix'N'Match, which provides a user-friendly framework for connecting identifier catalogues to Wikidata. {[}\protect\hyperlink{ref-JgiKEEdq}{150}/?p=114{]}, as seen in Figure \ref{fig:mixn_match_cl}.
Logically, we created a Mix'N'Match catalogue for harmonizing Cell Ontology IDs to Wikidata (\url{https://mix-n-match.toolforge.org/\#/catalog/4719}), harnessing the community support for the task.

>>>>

This manuscript
(\href{https://lubianat.github.io/quali_phd/v/52efe9a95b60519392be126b6111a6ac8f1ea5da/}{permalink})
was automatically generated
from \href{https://github.com/lubianat/quali_phd/tree/52efe9a95b60519392be126b6111a6ac8f1ea5da}{lubianat/quali\_phd@52efe9a}
on December 11, 2021.

@@

Human Cell Atlas publications {[}\protect\hyperlink{ref-2A7P3sTH}{124}{]} were selected and abstracts were annotated as richly as possible with Wikidata IDs using the hypothes.is annotation system (https://web.hypothes.is/). One article {[}\protect\hyperlink{ref-1GmbExweg}{2}{]}, describing the complete Human Cell Atlas project, was annotated in full. Annotations were retrieved via the hypothes.is API and processed with custom Python and R scripts (\url{https://github.com/lubianat/ann/tree/main/hypothesis_parsing}).

@@

As part of the research-action process, I have joined the Cell Ontology working group.
I participate in the monthly meetings and sporadic workshops, learning and contributing to the discussions.
Additionally, I contribute to the ontology development, actively engaging in the Cell Ontology GitHub repository (\url{https://github.com/obophenotype/cell-ontology}) and contributing with new terms and assertions.
I edit the ontology with the software for ontology editing Protégé v. 5.5.0 {[}\protect\hyperlink{ref-CHsAoEvr}{125}{]}.

@@

We targeted the question: which cell type definition allows crafting coherent biological statements?
The goal was not to say what cell types \emph{are}, but what they can be for a consistent representation on an ontology or a knowledge graph, like Wikidata.
We avoided the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{126}{]} for an example).
Even though such a distinction is essential for theoretical research, it is not required to represent biomedical experiments formally.

In a preprint derived from this PhD project {[}\protect\hyperlink{ref-uCE45Jh8}{127}{]}, we proposed naming conventions for different cell types classes.
Much of the literature mixes cell types in one species (e.g., when dealing with a cell type as an evolutionary unit) or multiple species (e.g., in the Cell Ontology).
It is helpful to distil these different concepts into names. Given the importance of the species' concept in biological classification {[}\protect\hyperlink{ref-AHGRV6KQ}{128}{]}, we derive a species-centric view on naming classes of cell types.
The four classes (Figure \ref{fig:classes}) we propose are as follows:

@@

Refining the different concepts around the notion of ``cell type'' is important but will require decades before a reasonable consensus. Here we adopt a liberal view of cell type, defining, for our purposes, a cell type as any class of cells described by a domain expert with evidence of the reality of its instances.
The requirement of evidence of existence in the material world is based on the principle of instantiation of ontological realism {[}\protect\hyperlink{ref-5ZFRyTy5}{129}{]}. Barry Smith and Werner Ceuster state ``A term should be included in a reference ontology only if there is experimental evidence that instances to which that term refers exist in reality. (`Exists' here should be understood in a tenseless sense in order to accommodate, for example, universals pertaining to extinct species as well as universals such as swarm or hurricane which are instantiated only intermittently.)''
Thus, in this work, one minimum requirement for a cell type to be catalogued is a public description of the class, with evidence for the existence of instances of the class in reality.

By ``class'', we mean an abstract entity in a sense intended by the multilevel theory (MLT) of conceptual modelling {[}\protect\hyperlink{ref-eP47FILn}{130}{]}
Figure \ref{fig:multilevel} displays a simplified version of MLT adopted throughout this project.
In this framework, real-world entities are materializations of \emph{individuals}.
\emph{Individuals} are theoretical constructs that are (1) thought to exist or have existed, as per the principle of instantiation, and (2) refer to only one (01) material entity at any point in time.
For example, Wikidata has entries for people, e.g.~``Helder Nakaya (\href{https://www.wikidata.org/wiki/Q42614737}{Q42614737})'' and ``Charles Darwin (\href{https://www.wikidata.org/wiki/Q42614737}{Q1035})'' are considered \emph{individuals} by Multi-Level Theory.
Other examples of \emph{individuals} include ``Albert Einstein's brain (\href{https://www.wikidata.org/wiki/Q2464312}{Q2464312})'' and the ``Christ the Redeemer statue (\href{https://www.wikidata.org/wiki/Q79961}{Q79961})''.

@@

For the practical purpose adopted here, we avoid the dissection of the differences between persistent classes of cells (often called ``cell types'') or the transient, fugacious classes of cells (often called ``cell states'') (see ``Definition of cell identity'' section in {[}\protect\hyperlink{ref-pFijpXkl}{126}{]} for an example). We also consider only the cell as it was observed in an experiment, not necessarily the future conditions of any cell (i.e., the ``cell fate''). {[}\protect\hyperlink{ref-OjxQDZpx}{31}{]}\\
Even though such a distinction is an important topic for theoretical research, it is outside the initial scope of this work.

@@

We also note that we made a judgment call to use the term ``cell type'' to emphasize the focus on types as classes (or ``kinds'') in contrast to real-world objects.
The term ``cell class'' is also used in the literature and is a suitable synonym for our notion of cell type.
We opted to frame our work around the term ``cell type'' due to its historical usage and familiarity for the life sciences community.
Other related terms a ``cell set,'' ``cell population,'' and ``cell cluster,'' can also reminisce of a specific, countable group of cells, frequently from the same experiment.
The term ``cell identity'' has also been suggested for avoiding the cell type/cell state dilemma {[}\protect\hyperlink{ref-WKbly37M}{42}{]}, but we avoid it to emphasize a nominalistic perspective (in the Popperian sense{[}\protect\hyperlink{ref-1C6LI68h6}{131}{]}).
In doing so, we reinforce the intent to represent the cell types \_reported to exist\_instead of stating bluntly which cell types \emph{exist} or are \emph{essential} for human beings.

@@

Making the Human Cell Atlas more useful via Wikidata also includes the connection of related databases.
PanglaoDB {[}\protect\hyperlink{ref-M0lqRYJb}{132}{]} {[}\protect\hyperlink{ref-T3PlZ3Vh}{133}{]} is a publically-available database that contains data and metadata on hundreds of single-cell RNA sequencing experiments.
It provides extensive information on cell types, genes, tissues, and cell type markers, obtained via automatic and manual methods.
It also displays a rich web user interface for easy data acquisition, including database dumps for bulk downloads.

As of 8 December June 2021, the article describing PanglaoDB had been cited 230 times.
Despite its use by the community, the database is on a 3-star category for Linked Open Data {[}\protect\hyperlink{ref-OGXYtGZ8}{134}{]} as it does not use the open semantic standards from W3C (RDF and SPARQL) needed for a 4-star rank, neither the links to external data via standard identifiers that make datasets 5-star.
Improving the data format toward W3C's gold standards is a valuable step in making biological knowledge FAIR (Findable, Accessible, Interoperable, and Reusable).
Thus, we aimed to provide a case study of making the core information of PandlaoDB available in a 5-star Linked Open Data Format while improving the modelling of the necessary concepts on Wikidata.

As of August 2020, Wikidata had 264 items categorized as a ``cell type'', considerably less than in specialized cell catalogues, which count over two thousand cell types {[}\protect\hyperlink{ref-agnqfdk6}{49},\protect\hyperlink{ref-4AEy2xhQ}{135}{]}.
Strikingly, there were also 23 items categorized as instances of ``cell (Q7868)''.
This classification is imprecise, as an instance of cell would be an individual named cell from a single named individual, an example of conceptual disarray that often occurs on Wikidata. {[}\protect\hyperlink{ref-imSZTi52}{136}{]}

@@

Now that we re-formatted the markers on PanglaoDB as Linked Open Data, we can make queries that were not possible before, including federated queries with other biological databases, such as Uniprot {[}\protect\hyperlink{ref-HUJ9NzAb}{137}{]}
and Wikipathways {[}\protect\hyperlink{ref-AgOOCR24}{138}{]}.
Due to previous similar reconciliation projects, Wikidata already contains information about genes, including their relations to Gene Ontology (GO) terms.

@@

This work exemplifies the power of releasing Linked Open Data via Wikidata, and provides the biomedical community with the first semantically accessible, 5-star LOD dataset of cell markers, easily reachable from Wikidata's SPARQL Query Service (\url{https://query.wikidata.org/}).
Alongside other case studies of biomedical data integration to Wikidata (see {[}\protect\hyperlink{ref-mPoPwN77}{116}{]}, it contributes with tools and practices to serve as basis for contributors.)
The work also paves the way for reconciling of other databases for cell-type markers, such as CellMarker {[}\protect\hyperlink{ref-chGii6yw}{47}{]}, labome {[}\protect\hyperlink{ref-rhRRCtlA}{139}{]}, CellFinder {[}\protect\hyperlink{ref-4AEy2xhQ}{135}{]} and SHOGoiN/CELLPEDIA {[}\protect\hyperlink{ref-6uWWsiSq}{140}{]}) (if the owners give proper authorization).
The approach we took here can be applied to any knowledge set of public interest, providing a low-cost and low-barrier platform for sharing biocurated knowledge in gold-standard format.

@@

Reading scientific articles is an integral part of the routine of modern scientists.
Although several literature-management software are available {[}\protect\hyperlink{ref-SdbIYKUa}{141}{]}, the process of reading is mainly artisanal.
There are no standard guidelines on how to probe the literature organize notes for biomedical researchers.
Thus, while reading and studying is a core activity, there are few (if any) protocols for the efficient screening of scientific articles.

Other professional traditions have dealt with similar issues in the past.
Notetaking is vital to keep track of financial balances and avoid costly problems in accounting.
Double-entry bookkeeping was developed in the 13th century as a professional solution for notetaking in accounting where ``every entry to an account requires a corresponding and opposite entry to a different account.'' {[}\protect\hyperlink{ref-uYuz0opI}{24}, =Double-entry\_bookkeeping\&oldid=1055066428{]}
In software development, Test-Driven Development (TDD) is a popular methodology where tests for code snippets are written before the code itself, therefore ensuring that written software passes minimum quality standards.
The similarities of Double-entry bookkeeping and TDD are diverse {[}\protect\hyperlink{ref-KNGv5MPp}{142}{]}, but for our purpose, here suffices to see both as professionalized systems that promote better quality and accountability of works.

In the humanities, there is a well-established practice of annotations of readings.
The annotation skills are part of standard academic training in the humanities {[}\protect\hyperlink{ref-rPKBwmYh}{143}{]}{[}\protect\hyperlink{ref-url}{\textbf{url?}}{]}.
An influential work in presenting methods for academic reading in the humanities is Umberto Eco's book ``How to Write a Thesis'' {[}\protect\hyperlink{ref-1HBVPtZGp}{144}{]}, which outlines not only \emph{how} to annotate the literature that basis an academic thesis, but also \emph{why} to do so.
The book, written originally in 1977, is still influential today.
Still, its theoretical scope (roughly the humanities) and its date preceding the digital era limits the extent to which it applies to the biomedical sciences.

@@

The \texttt{docs/} directory contains the live dashboard from the readings, which will be discussed in the following sessions.
The \texttt{downloads/} directory hosts the pdfs of the articles read with the system.
These are not committed to the repository and are only stored locally.
The \texttt{notes/} directory contains markdown files, one for each article read.
The \texttt{src/} directory contains the python code with the system's mechanics.
They contain helper functions for the command line commands discussed below:
- \texttt{wread} which receives a Wikidata QID for an article and outputs (1) a notes document, (2) a pdf for the paper obtained from Unpaywall {[}\protect\hyperlink{ref-15luL9zZC}{145}{]} and (3) an updated version of the dashboard HTML files in the \texttt{docs/} directory.
- \texttt{pop}, which ``pops'' an article from \texttt{toread.md} and runs \texttt{wread} for it
- \texttt{wadd}, which takes an URL for a Wikidata SPARQL query and adds new QIDs to \texttt{toread.md}
- \texttt{wadd\_all}, which parses \texttt{config.yaml} for recurrent SPARQL queries and runs \texttt{wadd} for each
- \texttt{wlog}, which adds, commits and pushes recent readings and dashboard updates to GitHub

All the structures described so far are commonly shared by any user of Wikidata Bib.
To personalize the use of the system, the user edits three plain text files.
\texttt{toread.md} hosts plain text QIDs of the articles that will be read.
These can be added either manually or via wadd.
While the \texttt{pop} command only sees QIDs, articles titles or other identifiers can temporarily be added to \texttt{toread.md} without breaking the system.
\texttt{index.md} hosts a numbered list of topics of interest.
This file plays the role of Umberto Eco's work plan, with the topics of interest for the academic. {[}\protect\hyperlink{ref-1HBVPtZGp}{144}{]}
These are used to tag articles for retrieval in a later step.
\texttt{config.yaml} contains shortcuts for different reading lists.
This is better explained by example.
In my \texttt{to\ read.md} file there are two reading lists, one following a \texttt{\#\ Cell\ types} header and another following a \texttt{\#\ Biocuration} header.
My \texttt{config.yaml} contains the following snippet:

@@

The Wikidata bib framework is coupled with a discipline of daily reading.
The discipline is inspired by Robert Cecil Martin's description of Test Driven Development in the book ``Clean Code'', which includes not only a technical description but a \emph{school of thought} of how software development might be approached. {[}\protect\hyperlink{ref-13HqB23xH}{146}{]}
Every day, I read one article of each list, using the notetaking station displayed in Figure {[}\protect\hyperlink{ref-fig}{\textbf{fig?}}{]}: notetaking.
The constancy of reading allows steady coverage of the relevant literature.
While the discipline has worked for this research project, it is not required to use the Wikidata Bib system.

@@

The discipline also includes, whenever possible, an improvement of the metadata about the article on Wikidata.
In {[}\protect\hyperlink{ref-fig}{\textbf{fig?}}{]}: notetaking B are shown the links included in the dashboard.
A link to a Scholia {[}\protect\hyperlink{ref-hxzL9pmm}{106}{]} profile allows identification of related articles from a series of pre-made SPARQL queries probing bibliography data on Wikidata.
While Scholia provides an overview of a given article, it does not allow direct curation of the metadata.
For that, two links are provided, one to Wikidata and one to Author Disambiguator {[}\protect\hyperlink{ref-1A9RvszKC}{147}{]}.
By accessing the Wikidata page for the entity, one can add new triples, for example, curating authors and topics of the article, which are then used by Scholia and by Wikidata Bib's dashboard.
Author Disambiguator is a wrapper of an Wikimedia API that facilitates disambiguating author names to unique identifiers on Wikidata, thus feeding the public knowledge graph of publication and authors.\\
Finally, a link to the article's DOI or full-text URL is provided and serves as a fallback when the automatic download fails.
Of note, while the metadata curation has a technical benefit to Wikidata and the dashboard, it also plays a theoretical role.
By curating metadata on authors, the user of Wikidata Bib can better understand the people they read, and expand their metascientific perspective on their domain of interest.

@@

To organize the code for rendering the dashboard, we created a python package, wbib, and deposited it in PyPi, making it available via \texttt{pip}. {[}\protect\hyperlink{ref-6chnW6cc}{148}{]}.
The package implements the logic for rendering complex Wikidata-based academic dashboards and is available in GitHub at \url{https://github.com/lubianat/wbib}.
It allows the user to build dashboards based on Wikidata records of information such as gender of authors, the region of author's institutions, topics of articles and similar metascientific information.
The dashboard is composed of SPARQL queries written for the Wikidata Query Service {[}\protect\hyperlink{ref-2wDsXBwd}{85}{]}
It also allows users to feed an arbitrary list of articles and obtain a custom dashboard.
Wikidata Bib obtains the HTML dashboards after feeding wbib the lists of articles read in total (\texttt{index.html}) or in pre-determined time spans (\texttt{last\_day.html}, \texttt{past\_week.html} and \texttt{past\_month.html} )

\begin{figure}
\hypertarget{fig:dashboard}{%
\centering
\includegraphics[width=0.85\textwidth,height=\textheight]{images/wikidata_bib_display.png}
\caption{Wikidata Bib queries for institutions of authors and most-read venues}\label{fig:dashboard}
}
\end{figure}

The dashboard includes not only a basic list of read articles, but also statistics on most read authors and most-read venues.
It also displays an interactive map of the institutions of articles read, permitting a glance at geographic biases in activities.
An example of queries is shown in \ref{fig:dashboard}.
As the queries are rendered live, they evolve in quality with the growth of Wikidata.
Finally, the clean 5-star-open data format enables users to adapt the queries to include different aspects of Wikidata.
For example, table \ref{tbl:articles_read_hca} showcases 10 articles that (1) I have read in the past year and (2) were authored by a speaker of the 1st Human Cell Atlas Latin America Single Cell RNA-seqData Analysis Workshop {[}\protect\hyperlink{ref-1hag8XE6}{149}{]}.
One practical application that the dashboard enables, thus, is to identify people in an event, institution or location that the user has read before, therefore catalyzing the possibility of collaborations.
Anecdotally, this strategy was tested successfully at Biohackathon Europe 2021 {[}\protect\hyperlink{ref-kHL3NVxk}{150}{]}, where I used the system both to identify possible collaborators and as a conversation starter.

@@

To start the improvement in the interplay of both databases, we proposed and got the approval of a specific Wikidata identifier for the Cell Ontology, the ``Cell Ontology ID'' (\url{https://www.wikidata.org/wiki/Property:P7963}).
IDs can be added to Wikidata entities and connected them to external databases enabling integrative SPARQL queries.
Besides using the common Wikidata interface, one can crowd-curate identifiers via a 3rd-party service, Mix'N'Match, which provides a user-friendly framework for connecting identifier catalogues to Wikidata. {[}\protect\hyperlink{ref-JgiKEEdq}{151}/?p=114{]}, as seen in Figure \ref{fig:mixn_match_cl}.
Logically, we created a Mix'N'Match catalogue for harmonizing Cell Ontology IDs to Wikidata (\url{https://mix-n-match.toolforge.org/\#/catalog/4719}), harnessing the community support for the task.
